---
title: "Reproduced analysis of FARS data"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.height = 5,
                      fig.width = 8)
```

###Results   


```{r}
library(tibble)
library(dplyr)
library(tidyr)
library(readr)
library(maps)
library(purrr)
library(knitr)
library(ggplot2)
library(scales)
library(ggthemes)
```


```{r}
load("../data/clean_fars.RData")
source("../R/fars_functions.R")
```


**Percentages of drivers testing positive by drug type, sex, and year group.**   
```{r}
clean_fars %>%
  mutate(year_cat = cut(year, breaks = c(1999, 2002, 2006, 2010),
                        labels = c("1999-2002", "2003-2006",
                                   "2007-2010"),
                        include.lowest = TRUE, right = TRUE)) %>%
  filter(!is.na(sex)) %>%
  group_by(drug_type, sex, year_cat) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1)) %>%
  select(drug_type, sex, year_cat, perc_positive) %>%
  unite(sex_year_cat, sex, year_cat) %>%
  spread(sex_year_cat, perc_positive) %>%
  kable(col.names = c("Drug type", "F 1999-2002",
                      "F 2003-2006", "F 2007-2010",
                      "M 1999-2002", "M 2003-2006",
                      "M 2007-2010"))

``` 
   
   
**Figure 1: Prevalence of nonalcoholic drugs in fatally injured drivers by year and age group**

```{r}
clean_fars %>%
  filter(drug_type != "Alcohol",
         !is.na(agecat))%>%
  group_by(unique_id, agecat, year) %>%
  summarize(positive_test = sum(positive_for_drug, na.rm = TRUE),
            positive = any(positive_test > 0)) %>% 
  ungroup() %>% 
  group_by(year, agecat) %>% 
  summarize(total_test = length(positive),
            positive = sum(positive),
            perc_positive = round(100 * positive / total_test , 1)) %>%

   
ggplot(aes(year, perc_positive, position = agecat))+
  geom_line()+
  theme_few()+
 geom_point(aes(shape = agecat))+
  scale_shape_manual("Age",values = c(17,16,15, 3))+
  labs (x="Year",
      y="Positive for Nonalcoholic Drugs %")+
  scale_y_continuous(limits = c(0,35))

```
     
**Figure 2: Prevalence of nonalcohol drugs in fatally injured drivers by year and drug type**   
```{r}
clean_fars %>%
  group_by(year, drug_type) %>% 
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1)) %>%
  filter(drug_type != "Alcohol") %>%
   
ggplot(aes(year, perc_positive, position = drug_type)) +
  geom_point(aes(shape = drug_type))+
  scale_shape_manual("Drug Type",values = c(17,16,15, 3, 7))+
  geom_line()+
  theme_few()+
  labs (x="Year",
      y="Positive for Drugs %")+
  scale_y_continuous(breaks= c(0, 4, 8,  12))
      
     
```
 
**Figure 3: Prevelance of cannabinoid drugs in fatally injured drivers by year and age group**       

```{r}
clean_fars %>%
  group_by(agecat, year, drug_type) %>% 
  
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1)) %>%
 filter(drug_type == "Cannabinoid") %>% 
  filter(!is.na(agecat)) %>% 
   
ggplot(aes(year, perc_positive, position = agecat))+
  geom_line()+
  geom_point(aes(shape = agecat))+
  scale_shape_manual("Age",values = c(17,16,15, 3))+
  theme_few()+
  labs (x="Year",
      y="Positive for Cannabinoid %")


```


   
**Prevalence of drugs in fatally injured drivers for 1999 and 2010 by drug type**   
```{r}
first_prevalence <- clean_fars %>%
      filter(year %in% c("1999", "2010")) %>%
      group_by(drug_type, year) %>%
      summarize(positive = sum(positive_for_drug, na.rm = TRUE),
                trials = sum(!is.na(positive_for_drug)))

second_prevalence <- data.frame(prevalence = perc_cis(x = first_prevalence$positive, 
                                             n = first_prevalence$trials))

first_prevalence<- tibble::rowid_to_column(as.data.frame(first_prevalence, var = "rowid"))
second_prevalence <- tibble::rowid_to_column(as.data.frame(second_prevalence, var = "rowid"))

prev_final <- first_prevalence %>% 
  full_join(second_prevalence, by = "rowid") %>% 
  select(drug_type, year, prevalence) %>% 
  spread(key = year, value = prevalence) %>% 
  rename(`Drug type` = drug_type)
kable(prev_final)
```
   
   
      
 **Statistics for testing for trend in prevalence of drugs over study years by drug type using Cochran-Armitage trend test**


```{r}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
               "Stimulant", "Cannabinoid", "Other")
drug_trend_tests_ca <- lapply(drug_list, test_trend_ca)
drug_trend_tests_ca <- dplyr::bind_rows(drug_trend_tests_ca) %>%
  mutate(drug = drug_list) %>%
  select(drug, Z, p.value)

drug_trend_tests_ca %>% 
  kable()
```

**Statistics for testing for trend in prevalence of drugs over study years by drug type using Wald test of logistic regression coefficient for “year”**    


```{r}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
"Stimulant", "Cannabinoid", "Other")
drug_trend_tests_log_reg <- lapply(drug_list, test_trend_log_reg)
drug_trend_tests_log_reg <- bind_rows(drug_trend_tests_log_reg) %>%
mutate(drug = drug_list) %>%
select(drug, Z, p.value)

drug_trend_tests_log_reg %>% 
kable()
```

